---
title: "Angulate Tortoise Data Assignments"
author: "Muhammad Uzair Davids"
date: "2025-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r}
library(readxl)
library(ggplot2)
library(visreg) #  check whether patterns in the data are properly described by the model
```


## Load data
```{r}
tortoise <- read_excel("C:/Program Files/Git/STA5014Z/data/tortoise.data.xls")
```

# ASSIGNMENT 1
## QUESTION 1: Scatter plot
```{r}
# Select columns 5 to 11
tort <- tortoise[, 5:11]

# Generate scatterplot matrix
pairs(tort, main = "Scatterplot Matrix of Tortoise Measurements")
```
From the scatter plots we see that there are outliers present in the data. 

## QUESTION 2
Length seems to be least strongly correlated with weight.

```{r}
# Deleting some data points
tort2 <- tort # copy data frame to tort2
 tort2 <- tort2[tort2$`Mid width` < 600,] # only those rows where mid-width < 600
 tort2 <- tort2[tort2$`Rear width` < 200,]
 tort2 <- tort2[tort2$`Curved width` > 100,]
 tort2 <- tort2[tort2$`Gular length` < 100,]
 tort2 <- tort2[tort2$`Plastron length` < 180,]
 
# Plot again
pairs(tort2, main = "Adjusted Scatterplot Matrix of Tortoise Measurements")
```

## QUESTION 3
After removing the outliers, it seems that Gular length is the least strongly
correlated.

```{r}
tort_cor <- round(cor(tort), 2)
tort_cor

# Load necessary library
library(ggcorrplot)

# Compute Pearson's correlation matrix
cor_matrix <- cor(tort, use = "complete.obs", method = "pearson")

# Print correlation matrix
print(cor_matrix)

# Visualize correlation matrix
ggcorrplot(cor_matrix, method = "circle", type = "lower", lab = TRUE, title = "Pearson Correlation Matrix")

```

## QUESTION 4

From the above, we see that mid width is the least strongly correlated.

```{r}
 tort3 <- tortoise[tortoise$Sex == "M" | tortoise$Sex == "F", ]
 tort3$Sex <- factor(tort3$Sex)
 
range <-  range(tort3$Weight)
range
```

## QUESTION 5
Therefore, the range is 584g to 1657g.

```{r}
ggplot(tort3, aes(x = Sex, y = Weight, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Comparison of Weight Between Male and Female Tortoises",
       x = "Sex",
       y = "Weight") +
  theme_minimal()
```

The median weight of males is greater than females, and there is greater 
variance in males.

```{r}
# Convert Sex to a factor and ensure "M" comes first
tort3$Sex <- factor(tort3$Sex, levels = c("M", "F"))

t <- t.test(Weight ~ Sex, data = tort3, var.equal = FALSE)
t

ci_lower <- round(t$conf.int[1])
ci_upper <- round(t$conf.int[2])

# Extract confidence interval and round to nearest gram
if (t$estimate["mean in group M"] < t$estimate["mean in group F"]) {
  ci_lower <- -ci_lower
  ci_upper <- -ci_upper
}

# Print the confidence interval
cat("95% Confidence Interval for the Difference (Males - Females): [", ci_lower, ",", ci_upper, "] grams\n")
```
The confidence intervals are 88 and 179g, and the small p-value indicates 
strong evidence against the null hypothesis. 

```{r}
m1 <- lm(`Weight` ~ `Mid width`, data = tort3)
summary(m1)
```

The slope coefficient = -0.05 which indicates a weak negative relationship, but 
the p-value of 0.71 which is greater than 0.05 means that there is weak evidence
to support the alternative hypothesis. 

```{r}
m2 <- lm(`Weight` ~ `Length`, data = tort3)
summary(m2)
```

```{r}
# Predict the weight for a tortoise with length 200 mm, and get the confidence interval
prediction <- predict(m2, newdata = data.frame(Length = 200), 
                      interval = "confidence", level = 0.95)
print(prediction)


# Predict the weight for a tortoise with length 180 mm, and get the confidence interval
prediction2 <- predict(m2, newdata = data.frame(Length = 180), 
                       interval = "confidence", level = 0.95)
print(prediction2)
```

# ASSIGNMENT 2
## QUESTION 1: Multiple linear regression models
Fit the tortoise data to a multiple linear regression model.

```{r}
multi_m1 <- lm(`Weight` ~ `Length` + `Curved width`, data = tort3)
summary(multi_m1)
```

The above shows us that the regression coefficient estimates are partial 
regression coefficients.

## QUESTION 2

These coefficients tell us how much `Weight` changes with a change in each 
variable, assuming that we keep other variables constant.


## QUESTION 3

Find the median of `Length` and `Curved width` and predict the `Weight` of a 
tortoise with this `Length` and `Curved width`.

```{r}
# Length summary
summary(tort3$Length)

# Curved width summary
summary(tort3$`Curved width`)

# Predict with these values
y <- -739.2003 + 6.9667 * (192.0)  + 2.1193 * (216.0)
y
```

## QUESTION 4

Calculate the 95% CI for the effect of `Length` on `Weight`.


```{r}
confint(multi_m1)
```

## QUESTION 5
Check the model.

```{r}
# Partial plots
par(mfrow = c(1,2))
cex.lab <- 2.5
cex.axis <- 1.8
par(mar = c(5,6,1,0))

# Length plot
visreg(multi_m1, "Length", line = list(col = "blue"), 
       points = list(cex = 1, pch = 16, col = 1), 
       ylab = "", xlab = "Length", cex.lab = cex.lab, cex.axis = cex.axis)

mtext("Weight", side = 2, line = 3.5, cex = cex.lab)

par(mar = c(5,3,1,3))

# Curved width plot
visreg(multi_m1, `Curved width`, line = list(col = "red"), 
       points = list(cex = 1, pch = 16, col = 1), 
       xlab = "Curved width", yaxt = "n", 
       ylab = "", cex.lab = cex.lab, cex.axis = cex.axis)
```


